working title Compatibility system and stygma size of pollen recipient as main predictors of heterospecific pollen effect

### 2018-12-11 Manuscript_draft/Manuscript_draft.Rmd
adding Romina changes to Intro, still have to fix abstract and correct references from comments

In natural systems plant species normally coexist and share their floral visitors with other species @bascompte2003. This pollinator sharing from the plant perspective at the pre-pollination stage can be negative due to competition @pauw2013 or positive due to facilitation @carvalheiro2014. Once the floral visitor has arrived to the flower, pollen deposition on the stigma can take place and hence ovule fertilization. An increasing number of visits generally correlates with higher chances of fertilization @engel2003. However this is not always the case, among these possible flower visitors we find also nectar robbers and pollen thiefs @inouye1980 and the quality of pollen that is deposit on the stigma is also highly relevant to the pollination succes @aizen2007. Moreover, other less study issues in the pollination process are conspecific pollen loss and the arrival of foreign pollen which can have important detrimental effects on species fitness @morales2008 @ashman2013.

@@

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012 @ashman2013 @fang2013, being the generalist species the ones that receive greater loads of heterospecific pollen @fang2013. Surprisingly, this low ranges of heterospecific pollen have been shown to decrease fitness greatly @thomson1982. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species @thomson1982 @galen1989 @neiland1999 which also show a noteworthy fitness decrease. Although the effect of close related species is predicted to be greater @ashman2013 the presence of pollen of non related species on multiple species @arceo2016 and the higher chances to coexist with a species that has less niche overlap (Ref) make foreign pollen from far related species also an important subject of study in order to understand the importance of heterospecific pollen in natural systems. Notwithstanding, the effect of heterospecific pollen of far and close related species at community level remains to be explored beyond single pairwise interactions.

@@

Interestingly, incompatibility system seems to play an important role in foreign pollen effect where species that are self incompatible would have stronger barriers towards heterospecific pollen than self compatible species @ashman2013. The type of incompatibility, sporophytic or gametophytic is related with the place of pollen recognition where the former take place at the sitgma level and the latter occurs within the style, this last late acting pollen recognition mechanism is associated with greater negative effect @barrett1988. Remarkably, there is a great variability in mating systems across populations @whitehead2018 and therefore predict an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Moreover, other traits such as number of pollen grains per flower and number of ovules have been tradittionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REF). Selfer species would have a reduction of herkogamy (REF) and less pollen production per ovule (REF) which can be interpretated as a reduction of pollen expoorted into the community. Other morphological traits, like stigma size can be determinant for the total pollen quatity that a stigma can receive and therefore related to do that pollen size would also play an important role. Example with pollen here.

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. Moreover, variation in sampling effort have been shown to be determinant to characterize pollen transfer interactions @arceo2018. Although plant-pollinator network and pollen network studies can give a first picture of the importance of foreign pollen is necessary to address how its effect is shaped with both traits and relatedness of the species. For this reason, in this study we have created an artificial co-flowering community with 10 species belonging to three different families with different traits where we try to test the following questions: 1) Does heterospecific pollen reduce seed set, if so, 2) Does heterospecific pollen effect depend on any floral trait? 3) Does heterospecific pollen effect depend on the relatedness of the species.

>>>>

In most ecosystems, plant species normally coexist and share their floral visitors with other species @bascompte2003. From the plants’ perspective, this pollinator sharing can be positive due to facilitation @carvalheiro2014 or negative due to competition at the pre-pollination stage Pauw (2013). An increasing number of visits often correlates with higher chances of fertilization @engel2003. However this is not always the case, among these possible flower visitors there are also nectar robbers and pollen thieves @inouye1980.  Receiving both sufficient quantity and quality deposited on the stigma is thus highly relevant to the pollination success of the plant @aizen2007.

@@

By visiting many plant species, many pollinators are responsible for conspecific pollen loss and the transport of foreign pollen, both of which can have important detrimental effects on species fitness @morales2008;  @ashman2013; @arceo2016. Foreign pollen arrival can play an important role in plant species fitness but outcomes are variable and appear to be context dependent as there is not always a decrease in fitness @morales2008. Some of this variation is likely due to the enormous variability of foreign pollen transferred across systems ranging from between 0 and 75 percent but most studies report ranges of heterospecific pollen between 0 and 20 percent of the total pollen load @bartomeus2008 @montgomery2012; @ashman2013; @fang2013, yet even these relatively low amounts of heterospecific pollen transferred can decrease fitness greatly @thomson1982. While we now have some understanding of the impacts of heterospecific pollen quantity, we have less understanding of other factors that could be driving the variation in impacts upon fitness.  @ashman2013 postulated the first predictive framework that identifies a need to understand how plant traits might mediate heterospecific pollen effect, whereby mating system and pollen size were predicted to potentially mediate the impact of foreign pollen transfer on plant fitness. This concept is supported by other studies, such as @tong2016 that demonstrate an asymmetrical effect in 6 species of Pedicularis whereby the pollen of long styled species was able to grow the full length of the style on short styled species but not vice versa. While this suggests that the impacts of heterospecific pollen may differ among pollen donor and recipient, few studies have been conducted to ascertain whether this pattern is in fact a trend or to identify the extent to which other plant traits are critical to heterospecific pollen impacts.

@@

Given the large variability in mating systems across populations @whitehead2018, it is difficult to determine potential impacts upon HP transfer yet incompatibility system is another plant trait that appears to play an important role in foreign pollen effect whereby species that are self incompatible have stronger barriers to heterospecific pollen than self-compatible species @ashman2013. The type of incompatibility, (i.e. whether sporophytic or gametophytic) is related to the location of pollen recognition; sporophytic incompatibility relates to signaling at the stigma surface while gametophytic occurs within the style.  This later acting pollen recognition mechanism is associated with greater negative effect than sporophytic recognistion @barrett1988. Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Moreover, other traits such as number of pollen grains per flower and number of ovules have been traditionally associated with the type of incompatibility system where species with higher pollen ovule ratios are predicted to be xenogamous and species with low pollen ovule ratios autogamous (REF). Selfer species are known to have a reduction of herkogamy (REF) and less pollen production per ovule (REF) which can be interpretated as a reduction of pollen exported into the community. Other morphological traits, like stigma size can be determinant for the total pollen quatity that a stigma can receive and therefore related to do that pollen size would also play an important role. Example with pollen here.

**Paragraph 3** Maybe connect with paragrph above?

Species with similar traits are more closely related  XXXXXXXXX.  (Refs? Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang (2016) ).  Several studies predict that the impact of HP transfer is likely to be greater for closely related species (Ashman and Arceo-Gómez (2013)).  Few studies however, have focused on the impacts of heterospecific pollen of distantly related species Thomson et al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013) ; also cite studies from pollen transfer networks here such as ). Further, a majority of plant species are generalist and thus receive visits from multiple different pollinators.  Given these are generally the ones that receive greater loads of heterospecific pollen Fang and Huang (2013) and unrelated species are more likely to coexist  with other species due to less niche overlap (Ref),  understanding the role of foreign pollen from distantly related species thus deserves greater attention in understanding coexistence blah blahXXXXX refs.. Notwithstanding, the effect of heterospecific pollen of far and close related species at community level remains to be explored beyond single pairwise interactions.

@@

In this study we investigated how floral reproductive traits and relatedness mediate the impact of HP transfer by asking the following research questions : To what extent do (i) floral reproductive traits and (ii) relatedness, mediate the impacts of heterospecific pollen on seed set.  We do this by creating an artificial co-flowering community with 10 species belonging to three different families with different traits.

### 2018-12-11 Manuscript_draft/Manuscript_draft.tex
adding Romina changes to Intro, still have to fix abstract and correct references from comments

In natural systems plant species normally coexist and share their floral
visitors with other species Bascompte et al. (2003). This pollinator
sharing from the plant perspective at the pre-pollination stage can be
negative due to competition Pauw (2013) or positive due to facilitation
Carvalheiro et al. (2014). Once the floral visitor has arrived to the
flower, pollen deposition on the stigma can take place and hence ovule
fertilization. An increasing number of visits generally correlates with
higher chances of fertilization Engel and Irwin (2003). However this is
not always the case, among these possible flower visitors we find also
nectar robbers and pollen thiefs Inouye (1980) and the quality of pollen
that is deposit on the stigma is also highly relevant to the pollination
succes Aizen and Harder (2007). Moreover, other less study issues in the
pollination process are conspecific pollen loss and the arrival of
foreign pollen which can have important detrimental effects on species
fitness Morales and Traveset (2008) Ashman and Arceo-Gómez (2013).

@@

Recent studies have advanced in the ecological understanding of
heterospecific pollen effect Morales and Traveset (2008) Ashman and
Arceo-Gómez (2013) Arceo-Gómez and Ashman (2016). A general overview of
foreign pollen arrival is that it can play an important role on species
fitness but seems to be context dependent and not always produce a
decrease in fitness Morales and Traveset (2008). Part of this
unpredictability is due to the enormous variability of foreing pollen
transferred in nature, where levels between 0 and 75 percent are seen,
but most commonly values ranges between 0 and 20 percent of the total
pollen load Bartomeus et al. (2008) Montgomery and Rathcke (2012) Ashman
and Arceo-Gómez (2013) Fang and Huang (2013), being the generalist
species the ones that receive greater loads of heterospecific pollen
Fang and Huang (2013). Surprisingly, this low ranges of heterospecific
pollen have been shown to decrease fitness greatly Thomson et al.
(1982). Although heterospecific pollen quantity is fundamental to
understand the outcome of the interaction so is the different traits of
both pollen donor and recipient. Ashman and Arceo-Gómez (2013)
postulated the first predictive framework for traits of heterospecific
pollen effect, where different traits such as compatibility system and
pollen size among others seems to be crucial to understand foreing
pollen effect. Moreover, in Tong and Huang (2016) an assymetric effect
was shown in a crossing experiment between 6 species of the genus
\emph{Pedicularis} where the pollen of long styled species was able to
grow the full length of the style on short styled species but not
viceversa. Despite these recent caveats, we still lack empirical
evidence to affirm what are the main traits that drive heterospecific
pollen effect for both pollen donor and recipient at seed production
level. Interestingly, to comprehend how these traits interact is also
crucial to look at the phylogenetic relatedness of the species. There is
a considerable amount of literature of crosses between close related
species Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and
Huang (2016) but few works focused on heterospecific pollen of far
related species Thomson et al. (1982) Galen and Gregory (1989) Neiland
and Wilcock (1999) which also show a noteworthy fitness decrease.
Although the effect of close related species is predicted to be greater
Ashman and Arceo-Gómez (2013) the presence of pollen of non related
species on multiple species Arceo-Gómez and Ashman (2016) and the higher
chances to coexist with a species that has less niche overlap (Ref) make
foreign pollen from far related species also an important subject of
study in order to understand the importance of heterospecific pollen in
natural systems. Notwithstanding, the effect of heterospecific pollen of
far and close related species at community level remains to be explored
beyond single pairwise interactions.

@@

Interestingly, incompatibility system seems to play an important role in
foreign pollen effect where species that are self incompatible would
have stronger barriers towards heterospecific pollen than self
compatible species Ashman and Arceo-Gómez (2013). The type of
incompatibility, sporophytic or gametophytic is related with the place
of pollen recognition where the former take place at the sitgma level
and the latter occurs within the style, this last late acting pollen
recognition mechanism is associated with greater negative effect Barrett
(1988). Remarkably, there is a great variability in mating systems
across populations Whitehead et al. (2018) and therefore predict an
effect of foreign pollen is a bit obscured by the variability within
species, however species that are strong selfers or strong outcrossers
have less variablity in mating systems and predictions of effect could
be more realistic (see figure 1 from Whitehead et al. (2018)). Moreover,
other traits such as number of pollen grains per flower and number of
ovules have been tradittionally associated with the type of
incompatibility system where species with higher pollen ovule ratios are
predicted to be xenogamous and species with low pollen ovule ratios
autogamous (REF). Selfer species would have a reduction of herkogamy
(REF) and less pollen production per ovule (REF) which can be
interpretated as a reduction of pollen expoorted into the community.
Other morphological traits, like stigma size can be determinant for the
total pollen quatity that a stigma can receive and therefore related to
do that pollen size would also play an important role. Example with
pollen here.

@@

The great environmental variability in natural systems and complexity of
floral structures make heterospecific pollination studies a daunting
task. Moreover, variation in sampling effort have been shown to be
determinant to characterize pollen transfer interactions Arceo-Gómez et
al. (2018). Although plant-pollinator network and pollen network studies
can give a first picture of the importance of foreign pollen is
necessary to address how its effect is shaped with both traits and
relatedness of the species. For this reason, in this study we have
created an artificial co-flowering community with 10 species belonging
to three different families with different traits where we try to test
the following questions: 1) Does heterospecific pollen reduce seed set,
if so, 2) Does heterospecific pollen effect depend on any floral trait?
3) Does heterospecific pollen effect depend on the relatedness of the
species.

@@

Results of hand cross pollination, self hand pollination, natural
selfing and apomixis are presented in \textbf{Table 2}. Heterospecific
pollen reduced seet set signifcatively with the 50-50\% heterospecific
pollen treatments for 65\% of the pairwise interactions
p\textless{}0.05. Across families we found a very similar effect but
when species where look at species level they respond differently even
within the same family, for instance for two species of the Brassicaceae
family \emph{Brassica oleracea} and \emph{Eruca versicaria} we found
very contrasting effects of foreign pollen where for the first one, all
donors reduce seed set significatively and for the second, just two
species did out of nine. The 100\% foreign pollen treatments barely
produced seeds or fruits and just for \emph{Sinapis alba} we did not
find significant differences between the hand cross pollination and one
treatment with pollen from a confamilial. Solanaceae species with berry
fruit type developed small fruits or even normal fruits in some cases.
\emph{S. lycopersicum} seems to produced small fruits (35\% of the
treatments) independently of pollen and pollen donor due to also
apomictic treatments did, never normal size. \emph{C. annum} produced
some fruits (9\%) of both small and normal size and finally \emph{S.
melongena} produced seedless normal fruits with just confamilial pollen
(3\%), for both species seems that fruit formation was induced by pollen
on the stigma because of no production of fruit in apomictic treatments.

@@

\hypertarget{ref-arceo2018}{}
Arceo-Gómez, G., C. Alonso, T.-L. Ashman, and V. Parra-Tabla. 2018.
Variation in sampling effort affects the observed richness of
plant--plant interactions via heterospecific pollen transfer:
Implications for interpretation of pollen transfer networks. American
journal of botany 105:1601--1608.

@@

\hypertarget{ref-arceo2016can}{}
Arceo-Gómez, G., R. A. Raguso, and M. A. Geber. 2016. Can plants evolve
tolerance mechanisms to heterospecific pollen effects? An experimental
test of the adaptive potential in clarkia species. Oikos 125:718--725.

@@

\hypertarget{ref-brown2001}{}
Brown, B. J., and R. J. Mitchell. 2001. Competition for pollination:
Effects of pollen of an invasive plant on seed set of a native congener.
Oecologia 129:43--49.

@@

\hypertarget{ref-galen1989}{}
Galen, C., and T. Gregory. 1989. Interspecific pollen transfer as a
mechanism of competition: Consequences of foreign pollen contamination
for seed set in the alpine wildflower, polemonium viscosum. Oecologia
81:120--123.

@@

\hypertarget{ref-neiland1999}{}
Neiland, M., and C. Wilcock. 1999. The presence of heterospecific pollen
on stigmas of nectariferous and nectarless orchids and its consequences
for their reproductive success. Protoplasma 208:65--75.

\hypertarget{ref-pauw2013}{}
Pauw, A. 2013. Can pollination niches facilitate plant coexistence?
Trends in ecology \& evolution 28:30--37.

>>>>

In most ecosystems, plant species normally coexist and share their
floral visitors with other species Bascompte et al. (2003). From the
plants' perspective, this pollinator sharing can be positive due to
facilitation Carvalheiro et al. (2014) or negative due to competition at
the pre-pollination stage Pauw (2013). An increasing number of visits
often correlates with higher chances of fertilization Engel and Irwin
(2003). However this is not always the case, among these possible flower
visitors there are also nectar robbers and pollen thieves Inouye (1980).
Receiving both sufficient quantity and quality deposited on the stigma
is thus highly relevant to the pollination success of the plant Aizen
and Harder (2007).

@@

By visiting many plant species, many pollinators are responsible for
conspecific pollen loss and the transport of foreign pollen, both of
which can have important detrimental effects on species fitness Morales
and Traveset (2008); Ashman and Arceo-Gómez (2013); Arceo-Gómez and
Ashman (2016). Foreign pollen arrival can play an important role in
plant species fitness but outcomes are variable and appear to be context
dependent as there is not always a decrease in fitness Morales and
Traveset (2008). Some of this variation is likely due to the enormous
variability of foreign pollen transferred across systems ranging from
between 0 and 75 percent but most studies report ranges of
heterospecific pollen between 0 and 20 percent of the total pollen load
Bartomeus et al. (2008) Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Fang and Huang (2013), yet even these relatively low
amounts of heterospecific pollen transferred can decrease fitness
greatly Thomson et al. (1982). While we now have some understanding of
the impacts of heterospecific pollen quantity, we have less
understanding of other factors that could be driving the variation in
impacts upon fitness. Ashman and Arceo-Gómez (2013) postulated the first
predictive framework that identifies a need to understand how plant
traits might mediate heterospecific pollen effect, whereby mating system
and pollen size were predicted to potentially mediate the impact of
foreign pollen transfer on plant fitness. This concept is supported by
other studies, such as Tong and Huang (2016) that demonstrate an
asymmetrical effect in 6 species of Pedicularis whereby the pollen of
long styled species was able to grow the full length of the style on
short styled species but not vice versa. While this suggests that the
impacts of heterospecific pollen may differ among pollen donor and
recipient, few studies have been conducted to ascertain whether this
pattern is in fact a trend or to identify the extent to which other
plant traits are critical to heterospecific pollen impacts.

@@

Given the large variability in mating systems across populations
Whitehead et al. (2018), it is difficult to determine potential impacts
upon HP transfer yet incompatibility system is another plant trait that
appears to play an important role in foreign pollen effect whereby
species that are self incompatible have stronger barriers to
heterospecific pollen than self-compatible species Ashman and
Arceo-Gómez (2013). The type of incompatibility, (i.e.~whether
sporophytic or gametophytic) is related to the location of pollen
recognition; sporophytic incompatibility relates to signaling at the
stigma surface while gametophytic occurs within the style. This later
acting pollen recognition mechanism is associated with greater negative
effect than sporophytic recognistion Barrett (1988). Nonetheless, an
effect of foreign pollen is a bit obscured by the variability within
species, however species that are strong selfers or strong outcrossers
have less variablity in mating systems and predictions of effect could
be more realistic (see figure 1 from Whitehead et al. (2018)). Moreover,
other traits such as number of pollen grains per flower and number of
ovules have been traditionally associated with the type of
incompatibility system where species with higher pollen ovule ratios are
predicted to be xenogamous and species with low pollen ovule ratios
autogamous (REF). Selfer species are known to have a reduction of
herkogamy (REF) and less pollen production per ovule (REF) which can be
interpretated as a reduction of pollen exported into the community.
Other morphological traits, like stigma size can be determinant for the
total pollen quatity that a stigma can receive and therefore related to
do that pollen size would also play an important role. Example with
pollen here.

\textbf{Paragraph 3} Maybe connect with paragrph above?

Species with similar traits are more closely related XXXXXXXXX. (Refs?
Brown and Mitchell (2001) Arceo-Gómez et al. (2016) Tong and Huang
(2016) ). Several studies predict that the impact of HP transfer is
likely to be greater for closely related species (Ashman and Arceo-Gómez
(2013)). Few studies however, have focused on the impacts of
heterospecific pollen of distantly related species Thomson et al. (1982)
Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects
and most stigmas have been found to carry multiple species of foreign
pollen with little attention to degree of relatedness (Arceo-Gómez and
Ashman (2016); Fang and Huang (2013) ; also cite studies from pollen
transfer networks here such as ). Further, a majority of plant species
are generalist and thus receive visits from multiple different
pollinators. Given these are generally the ones that receive greater
loads of heterospecific pollen Fang and Huang (2013) and unrelated
species are more likely to coexist with other species due to less niche
overlap (Ref), understanding the role of foreign pollen from distantly
related species thus deserves greater attention in understanding
coexistence blah blahXXXXX refs.. Notwithstanding, the effect of
heterospecific pollen of far and close related species at community
level remains to be explored beyond single pairwise interactions.

@@

In this study we investigated how floral reproductive traits and
relatedness mediate the impact of HP transfer by asking the following
research questions : To what extent do (i) floral reproductive traits
and (ii) relatedness, mediate the impacts of heterospecific pollen on
seed set. We do this by creating an artificial co-flowering community
with 10 species belonging to three different families with different
traits.

@@

Results of hand cross pollination, self hand pollination, natural
selfing and apomixis are presented in \textbf{Table 2}. Heterospecific
pollen reduced seet set signifcatively with the 50-50\% heterospecific
pollen treatments for 65\% of the pairwise interactions
p\textless{}0.05. Across families we found a very similar effect but
when species where look at species level they respond differently even
within the same family, for instance for two species of the Brassicaceae
family \emph{Brassica oleracea} and \emph{Eruca versicaria} we found
very contrasting effects of foreign pollen where for the first one, all
donors reduce seed set significatively and for the second, just two
species did out of nine. The 100\% foreign pollen treatments barely
produced seeds or fruits and just for \emph{Sinapis alba} we did not
find significant differences between the hand cross pollination and one
treatment with pollen from a confamilial. Solanaceae species with berry
fruit type developed small fruits or even normal fruits in some cases.
\emph{S. lycopersicum} seems to produced small fruits (35\% of the
treatments) independently of pollen and pollen donor due to also
apomictic treatments did, never normal size. \emph{C. annuum} produced
some fruits (9\%) of both small and normal size and finally \emph{S.
melongena} produced seedless normal fruits with just confamilial pollen
(3\%), for both species seems that fruit formation was induced by pollen
on the stigma because of lack of fruit production with apomictic
treatments.

@@

@@

@@

@@

@@
